---
title: "R Tutorial: Precipitation Graphs, Pt 3"
author: "Nathan Byers"
date: "Tuesday, July 24, 2014"
output: 
  html_document:
    fig_height: 6
    fig_width: 9
    theme: readable
---

In <a href="http://rpubs.com/NateByers/precipRidem" target="_blank">part 1</a>
of this tutorial we tidied up some data from the `IDEMdata` package, and in
<a href="http://rpubs.com/NateByers/precip2Ridem" target="_blank">part 2</a> we 
used that data to create precipitation graphs. In this part we will use functions
and loops to automate the creation of these graphs. 

## Functions

Functions in R are created using the following general format:

```{r, eval=FALSE}
myFunction <- function(){
  
}
```

So functions are created just like other variables in R, with the assignment
operator `<-`. Arguments for the function are placed inside the brackets `()` 
and the statements for the function are placed inside the braces `{}`. The last line
of the statements will be returned when the function is used. For example, here 
is a function that calculates the mean of a vector.

```{r}
myMean <- function(vector){
  sum.of.vector <- sum(vector)
  number.of.values <- length(vector)
  average <- sum.of.vector/number.of.values
  average
}
```

Now we can use this like any other function in R.

```{r}
myMean(vector = rnorm(10))
```

## Plot function

We're going to write our own function that will take watershed data and create
a precipitation graph. We'll call our function `prPlot()` and it'll have
the following general outline:

```{r, eval=FALSE}
prPlot <- function(date, precip, sample, trellis = FALSE, ...){
  
  if(trellis == FALSE){
    
    # lattice plot statements
    
  } else if(trellis == TRUE){
    
    # lattice plot statements with conditional
    
  } 
  
  # double y scale plot statements
  
}
  
```

The three arguments we have so far are the minimum arguments we need to create
a graph. We'll add more as we go along. Right now the requirements for these
arguments are that `date` has the `"Date"` class and that `precip` and `sample` have
the `"numeric"` class; that the three vectors are of the same length (and are
ordered appropriately); and that `trellis` is either `TRUE` or `FALSE`.

Now we'll simply add code for the plot into braces. 

```{r}
prPlot <- function(date, precip, sample, trellis, sample.label, ...){
  
  if(trellis == FALSE){
    
    rain <- xyplot(precip ~ date, type = "h", ylab = "Precipitation (in)",
                   xlab = "Date")

    sample <- xyplot(sample  ~ date, type = "p", ylab = sample.label, 
                     ylim = c(0, 2*max(sample, na.rm = TRUE)))
    
  } else if(trellis == TRUE){
    
    time <- as.POSIXlt(date)
    
    month.year <- format(time, "%b, %Y")
    
    month.year <- factor(month.year, levels = unique(month.year))
    
    rows <- length(levels(month.year))/3
    
    mday <- time$mday
    
    rain <- xyplot(precip ~ mday | month.year, type = "h",
                   ylab = "Precipitation (in)", layout = c(3, rows), as.table = TRUE, 
                   xlab = "")
    
    sample <- xyplot(sample  ~ mday | month.year, type = "p", ylab = sample.label,
                  ylim = c(0, 2*max(sample, na.rm = TRUE)))
    
  } 
  
  doubley <- doubleYScale(rain, sample, add.ylab2 = TRUE, under = TRUE, 
                          text = c( "Precipitation (in)", sample.label),
                          column = 2, type = c("l", "p"))
  
  update(doubley,
         par.settings = simpleTheme(col = c("black", "red"), pch = c(NA, 16)))
  
}
  
```

Before we can use this function, we need data in the proper format. So we'll write
a function that subsets the `wide.precip.df` data frame, properly formats the 
data, and calls the `prPlot` function.

```{r}
prSubsetPlot <- function(data, site, site.col, date.col, precip.col, 
                         sample.col, sample.name, conditioned = TRUE,
                         save = TRUE, file.prefix = NULL, 
                         file.extension = "png", ...){
  
  site.df <- data[data[, site.col] == site, c(date.col, sample.col)]
  
  precip.df <- unique(data[, c(date.col, precip.col)])
  
  site.df <- merge(precip.df, site.df, all = TRUE)
  
  remove(data, precip.df)
  
  plot <- prPlot(date = site.df[, date.col], precip = site.df[, precip.col],
                     sample = site.df[, sample.col], trellis = conditioned,
                     sample.label = sample.name)
  
  if(save){
    
    file.name <- paste(site, sample.col, "precip_graph", sep = "_")
    
    file.name <- paste(file.name, file.extension, sep = ".")
    
    file.name <- paste0(file.prefix, file.name)
    
    trellis.device(device = file.extension, filename = file.name)
    
    print(plot)
    
    dev.off()
    
  } else {
    
    plot
    
    }
  
}

```

This function takes a data frame (the argument `data`) that is in the wide format,
subsets it down to a particular sampling site, creates a precipitation graph
of the specified sampled data, and saves it. Let's see if it works. The following
code should save a `.png` file in your working directory.

```{r, eval=FALSE}
library(lattice)
library(latticeExtra)
library(IDEMdata)

data(wide_precip)

prSubsetPlot(data = wide.precip.df, site = "LMG-05-0009", site.col = "station", 
                 date.col = "date", precip.col = "precipitation",  sample.col = "sat", 
                 sample.name = "% Saturation", conditioned = "TRUE", save = TRUE, 
                 file.extension = "png")

```


## Loops

Like most programming languages, R has while loops and for loops. For loops are
the most common loops used in , so we'll cover how they are written. But in R, 
it's even more common to use a special family of loop operations called `apply()` 
functions. We'll end up using one of those to automate our precipitation graphs,
but first we'll go over the for loop.

## For loop

In most programming languages, the for loop would be set up in this way:

```{r, eval=FALSE}
i <- 1
for(i in sequence){
  # operation indexed by i
}
```

So you first create an index variable, traditionally named `i`, and give it an 
initial value of `1` (or `0`, or whatever initial value makes sense for the situation).
The `sequence` variable would be a sequence of values that you want the index 
variable to take while the statements in the brackets are iterated. Lets look at
a real example using the `myMean()` function we created above.

```{r}
list.of.vectors <- list(1:20, rnorm(10), rnorm(30))

i <- 1
vector.means <- c()
for(i in 1:3){
  vector.means[i] <- myMean(list.of.vectors[[i]])
  i <- i + 1
}

vector.means
```

So we created the index variable `i` with the intial value of `1`, created the
empty vector `vector.means` to collected the mean values in, and ran the for loop
over the sequence `1:3`. This means that in the first iteration the average of
the first vector in the list `list.of.vectors` is placed as the first value in
`vector.means`, i.e. we have the expression `vector.means[1] <- myMean(list.of.vectors[[1]])`.
And the last line of the loop adds a one to the index variable, so `i` takes on 
a value of `2` for the next iteration, and we have `vector.means[2] <- myMean(list.of.vectors[[2]])`.
This happens a third time, when the index value is `3`, and the for loop stops
after this iteration because `i` has taken the last value in the sequence.

In R, you can actually simplify the expression a little bit, because you don't 
have to manipulate the index variable. This will also work:

```{r}
vector.means <- c()
for(i in 1:3){
  vector.means[i] <- myMean(list.of.vectors[[i]])
}

vector.means
```

## apply functions

Experienced R programmers typically do not use for loops, especially for the 
example we just looked at. This is an ideal situation for the `lapply()` function.
`lappy()` is one of a family of functions that are referred to as `apply` functions.
Type `?apply` and `?lapply` to see descriptions of the many related functions.

`lapply()` takes just two arguments (at a minimum), `lapply(X, FUN, ...)`. `X`
is a list of objects that you want to perform an operation on, and `FUN` is the 
function that you want to apply to each object. In our situation, we want to 
apply `myMean()` to `list.of.vectors`.

```{r}
lapply(list.of.vectors, myMean)
```

`lapply()` always returns a list of objects. If we wanted the output to be a vector,
we could use the `sapply()` function which simplifies what is returned.

```{r}
sapply(list.of.vectors, myMean)
```

## Automating the graphs

Now, if we want to make a lot of precipitation graphs, we need to set up our code
so that we can easily loop through all of the arguments necessary for each plot.
For example, if we wanted to plot 3 different sampling substances at 2 sites, we
could visualize these scenarios in a data frame.

```{r}
substance <- c("sat", "e_coli", "ph")
station <- c("LMG-05-0009", "LMG-05-0004")
scenarios <- expand.grid(substance, station, stringsAsFactors = FALSE)
colnames(scenarios) <- c("substance", "station")
scenarios
```

One way to run the `precipSubsetPlot()` function on all of these scenarios is
to use a for loop that loops through the index of rows on the data frame. The only
thing we're missing is the label for the substance on the graph, so first we'll 
create a vector of substance names.

```{r, eval=FALSE}
substance.names <- scenarios[, "substance"]
substance.names[substance.names == "sat"] <- "% Saturation"
substance.names[substance.names == "e_coli"] <- "E. coli (MPN/100mL)"
substance.names[substance.names == "ph"] <- "pH (SU)"

for(i in 1:dim(scenarios)[1]){
  prSubsetPlot(data = wide.precip.df, site = scenarios[i, "station"], 
                   site.col = "station", date.col = "date", 
                   precip.col = "precipitation",  sample.col = scenarios[i, "substance"], 
                   sample.name = substance.names[i], conditioned = "TRUE", save = TRUE, 
                   file.extension = "png")
}
```

We could also use an `apply` function called `mapply()`.

```{r, eval=FALSE}
mapply(FUN = prSubsetPlot, site = scenarios[, "station"], 
       sample.col = scenarios[, "substance"], sample.name = substance.names,
       MoreArgs = list(data = wide.precip.df, site.col = "station", date.col = "date", 
                       precip.col = "precipitation", conditioned = "TRUE", save = TRUE, 
                       file.extension = "png"))
```

Finally, to put all of these funtions together and make the automation of 
precipitaiton graphs as simple as possible, we make one function that will loop
through all of the plots and save them to the working directory.

```{r, eval=FALSE}
precipPlot <- function(data, date.column = "date", sites, site.column = "station",
                       substances, substance.names, precip.column = "precipitation",
                       trellis = "TRUE", save = TRUE, file.extension = "png", ...){
  
  scenarios <- expand.grid(substances, sites, stringsAsFactors = FALSE)
  colnames(scenarios) <- c("substance", "site")
  
  mapply(FUN = prSubsetPlot, site = scenarios[, "site"], 
         sample.col = scenarios[, "substance"], sample.name = substance.names,
         MoreArgs = list(data = data, site.col = site.column, date.col = date.column,
                         precip.col = precip.column, conditioned = trellis,
                         save = save, file.extension = file.extension))

}

```

This function has been added to the `IDEMdata` package. To see the help file,
type `?precipPlot`. Below is how you would use this function to save a plot for
every substance and site in the `wide.precip.df` data frame.

```{r, eval=FALSE}

stations <- unique(wide.precip.df$station[!is.na(wide.precip.df$station)])
samples <- colnames(wide.precip.df)[8:34]
data(deep_river_chemistry)
library(reshape2)
wide.df <- dcast(deep_river_chemistry, STATION_NAME + ACTIVITY_NO + ACTIVITY_END_DATE 
                      + WATERBODY_NAME + UTM_EAST + UTM_NORTH + COUNTY_NAME 
                      ~ SUBSTANCE_NAME, value.var = 'LAB_RESULT')
names <- colnames(wide.df)[8:34]

precipPlot(data = wide.precip.df, sites = stations, substances = samples,
           substance.names = names)

```



...
-------------------------------------------------------------------------------


### R Markdown

This tutorial was created using RStudio's <a href="http://rmarkdown.rstudio.com/"
target="_blank">R Markdown</a>. The code can be found on <a href="https://github.com/NateByers/precip3Ridem"
target="_blank">GitHub</a>.







